It's COMPAS-licated -- Can We Find Fairness in the COMPAS Risk Assessment Algorithm?

### 2022-04-24 Paper_Draft1/Draft1.tex

%=================================================================
% Full title of the paper (Capitalized)
\Title{Does the COMPAS needle always point towards equity? Finding
Fairness in the COMPAS Risk Assessment Algorithm}

@@

The Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS) algorithm was created by the private for-profit company
Northpointe, also known by its parent company Equivant
\citep{equivant_faq_nodate}, to predict defendants' risk of recidivism.
It generates a score that classifies defendants' risk of recidivism as
either low, medium, or high \citep{angwin2016machine}. Jurisdictions
across the United States use the COMPAS risk assessment instrument. In
2016, ProPublica published a piece that analyzed the methods and
algorithms used by Northpointe to uncover racial biases in defendants'
scores \citep{angwin2016machine}. Their analysis looks at the
distribution of decile COMPAS scores among Black and white defendants.
ProPublica concludes, after using a statistical parity metric on false
positives or false negatives, that the algorithm is racially biased
\citep{larson2016we}. Northpointe denies the allegations of racial bias,
conducting their own analyses based on different statistical parity
metrics \citep{equivant_response_2018}. Since then, ProPublica has
responded, maintaining their original argument and Northpointe has yet
to respond (though in 2017, they adopted the name Equivant). Currently,
the two parties' disagreement focuses on pretrial Risk Assessment
Instruments (RAIs). ProPublica maintains that there are biases in the
outcome values, protected attributes, and covariates during
Northpointe's data processing phase. ProPublica accounts for these
biases in their analyses, and Northepointe's response, highlights how
ProPublica did not account for base rates of recidivism in their
analysis, which are important to understand initial percentages without
the presence of other information.

@@

\hypertarget{results}{%
\section{Results}\label{results}}

This section may be divided by subheadings. It should provide a concise
and precise description of the experimental results, their
interpretation as well as the experimental conclusions that can be
drawn.

\hypertarget{subsection-heading-here}{%
\subsection{Subsection Heading Here}\label{subsection-heading-here}}

Subsection text here.

\hypertarget{subsubsection-heading-here}{%
\subsubsection{Subsubsection Heading
Here}\label{subsubsection-heading-here}}

Bulleted lists look like this:

\begin{itemize}
\tightlist
\item
  First bullet
\item
  Second bullet
\item
  Third bullet
\end{itemize}

Numbered lists can be added as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  First item
\item
  Second item
\item
  Third item
\end{enumerate}

The text continues here.

All figures and tables should be cited in the main text as Figure
\ref{fig:mdpi-logo}, Table 1, etc.

{\centering \includegraphics{logo-mdpi}

\caption{This is a figure, Schemes follow the same formatting. If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel. (\textbf{b}) Description of what is contained in the second panel. Figures should be placed in the main text near to the first time they are cited. A caption on a single line should be centered.}\label{fig:mdpi-logo}
\end{figure}

Please see Table \ref{tab:mytable}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{x }\OtherTok{\textless{}{-}}\NormalTok{ tibble}\SpecialCharTok{::}\FunctionTok{tribble}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 1}\StringTok{\textasciigrave{}}\NormalTok{, }\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 2}\StringTok{\textasciigrave{}}\NormalTok{, }\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 3}\StringTok{\textasciigrave{}}\NormalTok{,}
\StringTok{"entry 1"}\NormalTok{, }\StringTok{"data"}\NormalTok{, }\StringTok{"data"}\NormalTok{,}
\StringTok{"entry 2"}\NormalTok{, }\StringTok{"data"}\NormalTok{, }\StringTok{"data"}
\NormalTok{)}

\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(x, }\AttributeTok{caption =} \StringTok{"This is a table caption. Tables should be placed in the main text near to the first time they are cited.}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{label\{tab:mytable\}"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[]{@{}lll@{}}
\caption{This is a table caption. Tables should be placed in the main
text near to the first time they are
cited.\label{tab:mytable}}\tabularnewline
\toprule
Title 1 & Title 2 & Title 3\tabularnewline
\midrule
\endfirsthead
\toprule
Title 1 & Title 2 & Title 3\tabularnewline
\midrule
\endhead
entry 1 & data & data\tabularnewline
entry 2 & data & data\tabularnewline
\bottomrule
\end{longtable}

This is an example of an equation:

\begin{equation}
\mathbb{S}
\end{equation}

Example of a theorem:

\begin{Theorem}
Example text of a theorem.
\end{Theorem}

The text continues here. Proofs must be formatted as follows:

Example of a proof:

\begin{proof}[Proof of Theorem 1]
Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
\end{proof}

The text continues here.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Authors should discuss the results and how they can be interpreted in
perspective of previous studies and of the working hypotheses. The
findings and their implications should be discussed in the broadest
context possible. Future research directions may also be highlighted.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

This section is not mandatory, but can be added to the manuscript if the
discussion is unusually long or complex.

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{It's COMPAS-licated -- Can We Find Fairness in the COMPAS Risk
Assessment Algorithm?}

@@

The Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS) algorithm was created by the private for-profit company
Northpointe, also known by its parent company Equivant
\citep{equivant_faq_nodate}, to predict defendants' risk of recidivism.
It generates a score that classifies defendants' risk of recidivism as
either low, medium, or high \citep{angwin2016machine}. Jurisdictions
across the United States use the COMPAS risk assessment instrument. In
2016, ProPublica published a piece that analyzed the methods and
algorithms used by Northpointe to uncover racial biases in defendants'
scores \citep{angwin2016machine}. Their analysis looks at the
distribution of decile COMPAS scores among Black and white defendants.
ProPublica concludes, after using a statistical parity metric on false
positives or false negatives, that the algorithm is racially biased
\citep{larson2016we}. Northpointe denies the allegations of racial bias,
conducting their own analyses based on different statistical parity
metrics \citep{equivant_response_2018}. Since then, the two parties have
had several exchanges, maintaining their original arguments. Currently,
the two parties' disagreement focuses on pretrial Risk Assessment
Instruments (RAIs). ProPublica maintains that there are biases in the
outcome values, protected attributes, and covariates during
Northpointe's data processing phase. ProPublica accounts for these
biases in their analyses, and Northepointe's response, highlights how
ProPublica did not account for base rates of recidivism in their
analysis, which are important to understand initial percentages without
the presence of other information.

@@
